# Password Philosophy

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

This is my attempt to solve [Day 2](https://adventofcode.com/2020/day/2).

```{r load data}
sample <- c("1-3 a: abcde",
            "1-3 b: cdefg",
            "2-9 c: ccccccccc")
actual <- read_lines("day_02_input.txt")
```

## Part 1

We need to convert each string into a list that contains the following:

-   minimum number of occurrences of the character
-   maximum number of occurrences of the character
-   the character that must be present
-   the password

Originally, I wrote the following function to handle this step:

```{r part 1 split input}
split_input <- function(input) {
  input %>%
    # first, split the string into the bit before the password, and the password
    str_split(": ") %>%
    # now, we can iterate over the results (each line of data)
    map(function(.x) {
      # find the min/max values from the first part of the input string .x
      mv <- .x[[1]] %>%
        # we can remove the last 2 characters of the string (the character)
        str_sub(1, -3) %>%
        # now we can split the string into the min and max values
        str_split("-") %>%
        # str_split is vectorised, but we only want the first results as we are 
        # dealing with one line at a time:
        # the first result will contain two values
        pluck(1) %>%
        # now convert these to integers
        as.integer()
      
      # we can now return the results
      list(
        min = mv[[1]],
        max = mv[[2]],
        character = str_sub(.x[[1]], -1, -1),
        password = .x[[2]]
      )
    })
}
```

But, after seeing a post by [\@antoine_fabri](https://twitter.com/antoine_fabri/status/1334197196795879426) on Twitter I discovered the `{unglue}` package which makes this much easier, and neater!

```{r part 1 split input (unglue)}
split_input <- function(input) {
  unglue::unglue(input, "{min}-{max} {character}: {password}", convert = TRUE)
}
```

We can see what our split_input function does to the sample data, looking at just the first item

```{r part 1 check split}
str(split_input(sample)[[1]])
```

Now we can build a function to return the valid passwords.

```{r part 1 validate sample passwords}
part_1_valid_passwords <- function(input) {
  is_valid <- function(i) {
    # extract the matching values from the string, we can then count the size of
    # the vector to find how many times that character appeared in the string
    lp <- length(str_extract_all(i$password, i$character)[[1]])
    # check to see if the password is valid
    i$min <= lp & lp <= i$max
  }
  
  input %>%
    split_input() %>%
    # run the is_valid function on each of the input values, keeping only the
    # valid passwords
    keep(is_valid) %>%
    # pull out the password value from each of the lists
    map_chr("password")
}
part_1_valid_passwords(sample)
```

This matches with the example given on the AOC website. We can now try to solve the first part:

```{r part 1 solution}
length(part_1_valid_passwords(actual))
```

## Part 2

We can use the `split_input` function again, but modifiy the `is_valid` function slightly.

```{r part 2 function}
part_2_valid_passwords <- function(input) {
  is_valid <- function(i) {
    # as per part 1, but this time extract the characters in the password at
    # position min and max
    a <- str_sub(i$password, i$min, i$min) == i$character
    b <- str_sub(i$password, i$max, i$max) == i$character
    # we can now xor these values: if both a and b are true xor = FASLE
    xor(a, b)
  }
  
  input %>%
    split_input() %>%
    keep(is_valid) %>%
    map_chr("password")
}
part_2_valid_passwords(sample)
```

This matches the example, so we can now solve part 2.

```{r part 2 solution}
length(part_2_valid_passwords(actual))
```
