# Crab Combat

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(crayon.enabled = NULL, # for when rendering on github actions
        scipen = 999)          # make sure we never print in scientific notation
library(tidyverse)
library(hash)
```

This is my attempt to solve [Day 22](https://adventofcode.com/2020/day/22).

```{r load data}
process_file <- function(file) {
  read_file(file) %>%
    str_trim() %>%
    str_remove_all("\r") %>%
    str_split("\n\n") %>%
    pluck(1) %>%
    map(~as.numeric(str_split(.x, "\n")[[1]][-1]))
}

sample <- process_file("samples/day_22_sample.txt")
actual <- process_file("inputs/day_22_input.txt")
```

## Part 1

We can create a function that applies the rules of the game.

```{r part 1 function}
part_1 <- function(input) {
  while (all(map(input, length) > 0)) {
    # get the first cards for player 1 and player 2
    a <- player_1[[1]]
    b <- player_2[[1]]
    
    # and remove the first cards
    player_1 <- player_1[-1]
    player_2 <- player_2[-1]
    
    if (a > b) {
      input[[1]] <- c(input[[1]], a, b)
    } else {
      input[[2]] <- c(input[[2]], b, a)
    }
  }
  
  # get the winners cards
  r <- if (length(input[[1]]) > 0) {
    input[[1]]
  } else {
    input[[2]]
  }
  
  # return the results as required for the puzzle
  sum(r * rev(seq_along(r)))
}
```

We can verify that this function returns the correct results against the sample data:

```{r part 1 sample test}
part_1(sample) == 306
```

And we can run the function against our actual data:

```{r part 1 actual}
part_1(actual)
```

## Part 2

Part 2 is a more complex problem. The naive implementation takes a number of minutes to run, but there is a useful
optimisation that
[u/daggerdraggon](https://www.reddit.com/r/adventofcode/comments/khyjgv/2020_day_22_solutions/ggpcsnd/) posted on
Reddit.

```{r}
part_2 <- function(input) {
  # create a function that can called recursively
  play <- function(player_1, player_2, subgame = FALSE) {
    prior_player_1 <- list()
    prior_player_2 <- list()
    
    # optimisation from u/daggerdragon on reddit
    # https://www.reddit.com/r/adventofcode/comments/khyjgv/2020_day_22_solutions/ggpcsnd/
    max_p1 <- max(player_1)
    max_p2 <- max(player_2)
    if (subgame &&
        max_p1 > max_p2 &&
        (length(player_1) + length(player_2)) < max_p1) {
      return (list(
          winner = 1,
          players = list(player_1, player_2)
        ))
    }
    
    # repeat while both players have cards
    while(length(player_1) > 0 & length(player_2) > 0) {
      # create key's to use in the hash of the prior hands seen hashtables
      player_1_str <- paste(player_1, collapse = ",")
      player_2_str <- paste(player_2, collapse = ",")
      
      # check, if we haven't seen this hand we will get NULL on lookups
      # looking up like this is orders of magnitude quicker than using %in%
      if (!is.null(prior_player_1[[player_1_str]]) &
          !is.null(prior_player_2[[player_2_str]])) {
        return (list(
          winner = 1,
          players = list(player_1, player_2)
        ))
      }
      
      # update the prior seen hashtables
      prior_player_1[[player_1_str]] <- TRUE
      prior_player_2[[player_2_str]] <- TRUE
      
      # get the first cards for player 1 and player 2
      a <- player_1[[1]]
      b <- player_2[[1]]
      
      # and remove the first cards
      player_1 <- player_1[-1]
      player_2 <- player_2[-1]
      
      # check to see if we can play a subgame or not
      r <- if (length(player_1) >= a & length(player_2) >= b) {
        # return the winner of the subgame
        play(player_1[1:a], player_2[1:b], TRUE)$winner
      } else if (a > b) {
        # player 1 wins
        1
      } else {
        # player 2 wins
        2
      }
      
      # update the player's hands if they have won
      if (r == 1) {
        player_1 <- c(player_1, a, b)
      } else {
        player_2 <- c(player_2, b, a)
      }
    }
    
    # we have run out of cards for one player, return the results
    list(
      winner = ifelse(length(player_1) > 0, 1, 2),
      players = list(player_1, player_2)
    )
  }
  
  # start to play the game
  p <- play(input[[1]], input[[2]])
  # get the results for the winner
  w <- p$players[[p$winner]]
  # return the results as required for the puzzle
  sum(w * rev(seq_along(w)))
}
```

We can test this function on the sample data:

```{r part 2 sample test}
part_2(sample) == 291
```

And we can run this function on the actual data:

```{r part 2 actual}
part_2(actual)
```
