# Handy Haversacks

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(crayon.enabled = NULL)
library(tidyverse)
library(igraph)
library(ggraph)
library(unglue)
```

This is my attempt to solve [Day 7](https://adventofcode.com/2020/day/7).

```{r load data}
sample <- read_lines("samples/day_07_sample.txt")
actual <- read_lines("inputs/day_07_input.txt")
```

## Part 1

I think today's problem is naturally solved with using graph's, so I'm going to use the `{igraph}` package. The easiest
way to create a graph is to first create a data frame that contains the columns `from` and `to` that indicate which
edges are in the graph. Any additional columns will be added as properties to the edge.

```{r part 1 convert input}
convert_input_to_tibble <- function(input) {
  input %>%
    # we don't need to keep the word's "bag" or "bags". In our data these words
    # always have a space before, and sometimes have a "." at the end. We can
    # use the following regex to remove these words
    str_remove_all(" bags?\\.?") %>%
    # we can now split our data at the word "contains"
    unglue_data("{from} contain {contains}") %>%
    # now we can split the contains column by comma's
    mutate(across(contains, str_split, ", ")) %>%
    # and expand the nested "contains" column
    unnest_longer(contains) %>%
    # handle edge case: if the string is no other need to insert a 0 at the start
    mutate(across(contains, ~if_else(.x == "no other", "0 no other", .x))) %>%
    # now we can split the contains column into the "n" part and the "to" part
    separate(contains, c("n", "to"), "(?<=\\d) ", convert = TRUE) %>%
    # reorder the columns
    select(from, to, n)
}
sample_df <- convert_input_to_tibble(sample)
actual_df <- convert_input_to_tibble(actual)
sample_df
```
We can create a graph using the `graph_from_data_frame()` function:

```{r create sample graph}
sample_g <- graph_from_data_frame(sample_df)
sample_g
```

and visualise this graph using `{igraph}`:

```{r show sample graph}
ggraph(sample_g, layout = "igraph", algorithm = "nicely") +
  geom_edge_link(aes(label = n), 
                 angle_calc = "along",
                 label_dodge = unit(2.5, "mm"),
                 arrow = arrow(length = unit(3, "mm")), 
                 end_cap = circle(10, "mm")) + 
  geom_node_label(aes(label = str_replace(name, " ", "\n")))
```

Our challenge is to find all of the vertices that have a path in to "shiny gold". The `subcomponent()` function can
tell us all of the vertices in a graph which reach a given vertex. We can simply take the length of this subcomponent
and subtract 1 (as the subcomponent includes "shiny gold").

```{r part 1 sample test}
length(subcomponent(sample_g, "shiny gold", "in")) - 1 == 4
```

We now just need to solve for the actual data.

```{r part 1 actual}
actual_g <- graph_from_data_frame(actual_df)
length(subcomponent(actual_g, "shiny gold", "in")) - 1
```

## Part 2

Part 2 sounds like a recursive function. We can first find the subgraph that includes all of the vertices from "shiny
gold". We can then convert this graph to an adjacency matrix and iterate through each vertex, recursively calling a
function that sums how many bags this bag will contain.

```{r part 2 function}
part_2 <- function(input) {
  sg <- induced_subgraph(input, subcomponent(input, "shiny gold", "out"))
  am <- as_adjacency_matrix(sg, attr = "n", sparse = FALSE)
  
  fn <- function(am, v = "shiny gold", n = 1) {
    a <- am[v, ] * n
    a <- a[a > 0] # keep as separate step, otherwise can end up with just scalar
    
    # now, iterate over each item in a and recursively call this function
    sum(map_dbl(names(a), ~fn(am, .x, a[[.x]]))) + sum(a)
  }
  
  # call this function on the base case
  fn(am)
}
```

We can test that our function works against the sample case:

```{r part 2 test sample}
part_2(sample_g) == 32
```

A second sample is provided, so we can test against that also:

```{r part 2 second sample test}
c("shiny gold bags contain 2 dark red bags.",
  "dark red bags contain 2 dark orange bags.",
  "dark orange bags contain 2 dark yellow bags.",
  "dark yellow bags contain 2 dark green bags.",
  "dark green bags contain 2 dark blue bags.",
  "dark blue bags contain 2 dark violet bags.",
  "dark violet bags contain no other bags.") %>%
  convert_input_to_tibble() %>%
  graph_from_data_frame() %>%
  part_2() == 126
```

Now we can run against the actual data:

```{r part 2 actual}
part_2(actual_g)
```

# Extra: alternative solution to part 2

We could also solve part 2 by iterating through each vertex and finding the incident edges and adjacent vertices. We
multiply the edge weight (`ie$n`) by the current value of `n`, sum these values and add `n` back in to the total. This
gives us the total number of bags, including the initial bag, so we need to subtract 1 from this answer.

```{r extra part 2 alt}
part_2_alt <- function(g, v, n) {
  ie <- incident_edges(g, v, "out")[[1]]
  vn <- ends(g, ie)[, 2]
  sum(map2_dbl(vn, ie$n * n, part_2_alt, g = g)) + n
}
```

We can now verify that this alternative function works as above.

```{r extra part 2 alt sample test}
part_2_alt(sample_g, "shiny gold", 1) - 1 == part_2(sample_g)
```

```{r extra part 2 alt actual test}
part_2_alt(actual_g, "shiny gold", 1) - 1 == part_2(actual_g)
```
