# Conway Cubes

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(crayon.enabled = NULL, # for when rendering on github actions
        scipen = 999)          # make sure we never print in scientific notation
library(tidyverse)
library(unglue)

START_TIME <- Sys.time()
```

This is my attempt to solve [Day 17](https://adventofcode.com/2020/day/17).

```{r load data}
sample <- read_lines("samples/day_17_sample.txt")
actual <- read_lines("inputs/day_17_input.txt")
```

## Part 1

We can use the R "array" data type for today's problem: these are n-dimensional vectors. First, let's create a function
to convert our input into a 3 dimensional array.

```{r part 1 to array}
to_array <- function(input, dimensions) {
  if (dimensions < 3) {
    stop("dimensions must be at least 3")
  }
  
  replacement <- c("." = 0, "#" = 1)

  m <- input %>%
    str_extract_all(".", simplify = TRUE) %>%
    apply(c(1,2), function(.x) replacement[[.x]])
  
  array(m, dim = c(nrow(m), ncol(m), rep(1, dimensions - 2)))
}

to_array(sample, 3)
```

At each iteration we need to increase the size of the state by 1 on every side.

```{r part 1 grow state}
grow_state_3d <- function(state) {
  ds <- dim(state)
  
  next_state <- array(0, dim = ds + 2)
  
  for (x in 1:ds[[1]]) {
    for (y in 1:ds[[2]]) {
      for (z in 1:ds[[3]]) {
        next_state[x + 1, y + 1, z + 1] <- state[x, y, z]
      }
    }
  }
  
  next_state
}
```

We need a function to iterate the states according to the rules.

```{r}
iterate_state_3d <- function(state) {
  state <- grow_state_3d(state)
  ds <- dim(state)
  next_state <- state
  
  # helper function to find the bounds of the array in a given dimension:
  # make sure we start no lower than 1, and no higher than the size of that dim
  gs <- function(a, i) max(1, a - 1):min(ds[[i]], a + 1)
 
  for (x in 1:ds[[1]]) {
    for (y in 1:ds[[2]]) {
      for (z in 1:ds[[3]]) {
        # get the bounds of the array for the dimensions
        xs <- gs(x, 1)
        ys <- gs(y, 2)
        zs <- gs(z, 3)
        # sum the neighbours, subtract the current cell
        e <- sum(state[xs, ys, zs]) - state[x, y, z]
        # apply rules
        if (state[x, y, z] == 1) {
          if (e != 2 & e != 3) {
            next_state[x, y, z] <- 0
          }
        } else {
          if (e == 3) {
            next_state[x, y, z] <- 1
          }
        }
      }
    }
  }
  # return the iterated state
  next_state 
}
```

We can build a helper function to run as many iterations as requested:

```{r part 1 run n iterations}
run_n_iterations_3d <- function(input, n) {
  reduce(1:n, ~iterate_state_3d(.x), .init = to_array(input, 3))
}
```

Now we can check our function runs as expected

```{r part 1 sample test}
sample %>%
  run_n_iterations_3d(6) %>%
  sum() == 112
```

And we can run with our actual data

```{r part 1 actual}
actual %>%
  run_n_iterations_3d(6) %>%
  sum()
```

## Part 2

For part 2 we just need to extend our functions to include a 4th dimension.

```{r}
grow_state_4d <- function(state) {
  ds <- dim(state)
  
  next_state <- array(0, dim = ds + 2)
  
  for (x in 1:ds[[1]]) {
    for (y in 1:ds[[2]]) {
      for (z in 1:ds[[3]]) {
        for (w in 1:ds[[4]]) {
          next_state[x + 1, y + 1, z + 1, w + 1] <- state[x, y, z, w]
        }
      }
    }
  }
  
  next_state
}

iterate_state_4d <- function(state) {
  state <- grow_state_4d(state)
  ds <- dim(state)
  next_state <- state
  
  gs <- function(a, i) max(1, a - 1):min(ds[[i]], a + 1)
 
  for (x in 1:ds[[1]]) {
    for (y in 1:ds[[2]]) {
      for (z in 1:ds[[3]]) {
        for (w in 1:ds[[4]]) {
          xs <- gs(x, 1)
          ys <- gs(y, 2)
          zs <- gs(z, 3)
          ws <- gs(w, 4)
          
          e <- sum(state[xs, ys, zs, ws]) - state[x, y, z, w]
          
          if (state[x, y, z, w] == 1) {
            if (e != 2 & e != 3) {
              next_state[x, y, z, w] <- 0
            }
          } else {
            if (e == 3) {
              next_state[x, y, z, w] <- 1
            }
          }
        }
      }
    }
  }
  
  next_state 
}

run_n_iterations_4d <- function(input, n) {
  reduce(1:n, ~iterate_state_4d(.x), .init = to_array(input, 4))
}
```

We can run with our sample data to test:

```{r part 2 sample test}
sample %>%
  run_n_iterations_4d(6) %>%
  sum() == 848
```

And we can run with our actual data:

```{r part 2 actual}
actual %>%
  run_n_iterations_4d(6) %>%
  sum()
```

---

*Elapsed Time: `r round(Sys.time() - START_TIME, 3)`s*
