# Operation Order

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(crayon.enabled = NULL, # for when rendering on github actions
        scipen = 999)          # make sure we never print in scientific notation
library(tidyverse)
library(unglue)
library(R6)
```

This is my attempt to solve [Day 18](https://adventofcode.com/2020/day/18).

```{r load data}
sample <- read_lines("samples/day_18_sample.txt")
actual <- read_lines("inputs/day_18_input.txt")
```

## Part 1

First I think it would be useful to have a [stack](https://en.wikipedia.org/wiki/Stack_(abstract_data_type)) to solve
today's problem. This is pretty simple to construct using R6. Here I use a fixed stack size (default to 100).

```{r stack class}
Stack <- R6Class(
  "Stack",
  public = list(
    initialize = function(stack_size = 100) {
      private$stack <- vector("list", stack_size)
      private$stack_size <- stack_size
    },
    push = function(v) {
      if (self$is_full()) {
        stop("stack is full")
      }
      private$ptr <- private$ptr + 1
      private$stack[[private$ptr]] <- v
    },
    pop = function() {
      if (self$is_empty()) {
        stop("no items in stack")
      }
      v <- private$stack[[private$ptr]]
      private$ptr <- private$ptr - 1
      v
    },
    is_empty = function() {
      private$ptr == 0
    },
    is_full = function() {
      private$ptr == private$stack_size
    }
  ),
  private = list(
    ptr = 0,
    stack_size = 0,
    stack = NULL
  )
)
```

We can then create a function which will step through the input and add items to the stack. When we a function and a
value we can update an accumulator (which we default to 0). If we reach a `(` we can push the accumulator and function
to the stack and reset these to the defaults. When we reach a `)` we can pop the values off the stack and start using
these again.

```{r part 1 function}
part_1 <- function(input) {
  part_1_parse <- function(input) {
    stack <- Stack$new()
    
    i <- 0
    acc <- 0
    fn <- `+`
    
    while (i < length(input)) {
      s <- input[[(i <- i + 1)]]
      if (s == "(") {
        stack$push(acc)
        stack$push(fn)
        
        acc <- 0
        fn <- `+`
        next()
      } else if (s == ")") {
        fn <- stack$pop()
        acc <- fn(stack$pop(), acc)
      } else if (s == "+" | s == "*") {
        fn <- get(s)
      } else {
        acc <- fn(acc, as.numeric(s))
      }
    }
    
    acc
  }
  
  input %>%
    str_replace_all(" ", "") %>%
    str_extract_all(".") %>%
    map_dbl(part_1_parse)
}
```

We can now check out function matches the provided sample:

```{r part 1 sample test}
all(part_1(sample) == c(71, 51, 26, 437, 12240, 13632))
```

And we can run the function on the actual data.

```{r part 1 actual}
sum(part_1(actual))
```

## Part 2

Part 2 can make use of our stack again, but this time we first parse the data by iterating over and adding all of the
values we would like to multiply to the stack. We immediately run addition and add these values to the stack instead.

Our parse function this time is recursive, when we reach a `(` we call our function again to parse that block. This 
returns the value of the accumulator and the position that we got up to.

```{r part 2 function}
part_2 <- function(input) {
  part_2_parse <- function(input, i = 1) {
    stack <- Stack$new()
    
    while (i <= length(input)) {
      a <- input[[i]]
      
      # we have found a bracket, escape loop
      if (a == ")") break()
      
      if (a == "(") {
        p <- part_2_parse(input, i + 1)
        # add value from inside bracket to stack
        stack$push(as.numeric(p[[1]]))
        # skip to matching parenthesis
        i <- p[[2]]
      } else if (a == "+") {
        i <- i + 1
        b <- input[[i]]
        if (b == "(") {
          p <- part_2_parse(input, i + 1)
          b <- as.numeric(p$acc)
          i <- p$i
        }
        av <- stack$pop()
        bv <- as.numeric(b)
        stack$push(av + bv)
      } else if (a == "*") {
        i <- i + 1
        b <- input[[i]]
        if (b == "(") {
          p <- part_2_parse(input, i + 1)
          b <- as.numeric(p$acc)
          i <- p$i
        }
        bv <- as.numeric(b)
        stack$push(bv)
      } else {
        stack$push(as.numeric(a))
      }
      
      i <- i + 1
    }
    
    acc <- 1
    while (!stack$is_empty()) {
      acc <- acc * stack$pop()
    }
    
    # return the accumulated value and the index we got to
    list(acc = acc, i = i)
  }
  
  input %>%
    str_replace_all(" ", "") %>%
    str_extract_all(".") %>%
    map(part_2_parse) %>%
    map_dbl("acc")
}
```

We can now check out function matches the provided sample:

```{r part 2 sample test}
all(part_2(sample) == c(231, 51, 46, 1445, 669060, 23340))
```

And we can run the function on the actual data.

```{r part 2 actual}
sum(part_2(actual))
```
