# Docking Data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(crayon.enabled = NULL, # for when rendering on github actions
        scipen = 999)          # make sure we never print in scientific notation
library(tidyverse)
library(unglue)
```

This is my attempt to solve [Day 14](https://adventofcode.com/2021/day/14).

```{r load data}
sample <- read_lines("samples/day_14_sample.txt")
actual <- read_lines("inputs/day_14_input.txt")
```

## Part 1

Part 1 requires us to apply a "bitmask" to a number. The function `packBits()` can convert a vector of `raw` values
back to an integer, but there are 2 issues.

First, integers in R are signed 32-bit values, and our mask is 36-bits long. We need to handle the sign bit separately,
otherwise we may end up with negative numbers.

Secondly, `packBits()` is big [endian](https://en.wikipedia.org/wiki/Endianness), but the mask is little endian. So we
need to reverse the values in the mask.

We can solve today's problem by taking two copies of the mask. `mask0` will be where we apply the replace 0 logic. We
can achieve this with a bitwise and, and a mask of all 1's except where the mask is a 0. `mask1` will be where we apply
the replace 1 logic. We can achieve this with a bitwise or, and a mask of all 0's except where the mask is a 1.

```{r part 1 apply mask}
part_1_apply_mask <- function(value, mask) {
  mask <- str_extract_all(mask, ".")[[1]] %>% rev()

  mask0 <- str_replace(mask, "X", "1") %>% as.raw()
  mask1 <- str_replace(mask, "X", "0") %>% as.raw()
  
  # ignore the most significant bit... assume all our numbers are positive...
  mask0_lsb <- packBits(c(mask0[1:31], raw(1)), "integer")
  mask1_lsb <- packBits(c(mask1[1:31], raw(1)), "integer")
  
  mask1_msb <- packBits(c(mask1[32:36], raw(27)), "integer")
  
  v <- value %>%
    bitwAnd(mask0_lsb) %>%
    bitwOr(mask1_lsb)
  
  as.numeric(v) + as.numeric(mask1_msb) * (2 ** 31)
}
```

We can now solve part 1 by building a data frame that has a column for the mask along with the value. Note that if a
memory location is updated twice we only keep the last value, so we can group by the memory locations and just keep
the last value.

```{r part 1 function}
part_1 <- function(input) {
  input %>%
    unglue_data("{instruction} = {value}") %>%
    mutate(mask = ifelse(instruction == "mask", value, NA)) %>%
    fill(mask) %>%
    filter(instruction != "mask") %>%
    mutate(across(instruction, str_extract, "\\d+"),
           across(c(instruction, value), as.integer)) %>%
    group_by(instruction) %>%
    summarise(across(everything(), last), .groups = "drop") %>%
    mutate(new_val = map2_dbl(value, mask, part_1_apply_mask)) %>%
    pull(new_val) %>%
    sum()
}
```

We can now test our function with the sample data:

```{r part 1 sample test}
part_1(sample) == 165
```

And we can run with the actual data:

```{r part 1 actual}
part_1(actual)
```

# Part 2

In part 2 we have a more complex scenario. We create a new "apply mask" function that will apply these rules. This
function will take the current memory, apply the rules for the current row of data, then return the memory. This can be
used in within `reduce()` to iterate over the data and update the memory.

```{r part 2}
part_2_apply_mask <- function(memory, mem_loc, value, mask) {
  m <- str_extract_all(mask, ".")[[1]]
  
  # convert mem_loc to a binary number
  mem_loc_b <- c(raw(4), rev(intToBits(mem_loc)))
  
  all_mem_locs <- list(mem_loc_b)
  
  for (i in 1:36) {
    if (m[[i]] == "1") {
      for (j in seq_along(all_mem_locs)) {
        all_mem_locs[[j]][[i]] <- as.raw(1)
      }
    } else if (m[[i]] == "X") {
      a <- all_mem_locs
      b <- all_mem_locs
      
      for (j in seq_along(all_mem_locs)) {
        a[[j]][[i]] <- as.raw(0)
        b[[j]][[i]] <- as.raw(1)
      }
      
      all_mem_locs <- c(a, b)
    }
  }
  
  for (ml in all_mem_locs) {
    mls <- paste(ml, collapse = "")
    memory[[mls]] <- value
  }
  
  memory
}
```

Part 2 provides a new sample

```{r part 2 sample}
sample_2 <- c("mask = 000000000000000000000000000000X1001X",
              "mem[42] = 100",
              "mask = 00000000000000000000000000000000X0XX",
              "mem[26] = 1")
```

We build a function now for solving part 2. We start by reusing the data processing step from part 1. We can then use
a little helper function to run the `part_2_apply_mask()` function, and put it all together with a `reduce()` over the
data.

```{r part 2 function}
part_2 <- function(input) {
  data <- input %>%
    unglue_data("{instruction} = {value}") %>%
    mutate(mask = ifelse(instruction == "mask", value, NA)) %>%
    fill(mask) %>%
    filter(instruction != "mask") %>%
    mutate(across(instruction, str_extract, "\\d+"),
           across(c(instruction, value), as.integer)) %>%
    rename(mem_loc = instruction)
  
  part_2_fn <- function(memory, data_row) {
    with(data_row, {
      part_2_apply_mask(memory, mem_loc, value, mask)
    })
  }
  
  data %>%
    transpose() %>%
    reduce(part_2_fn, .init = list()) %>%
    flatten_dbl() %>%
    sum()
}
```

We can now run our function on the new sample data

```{r part 2 sample test}
part_2(sample_2) == 208
```

```{r part 2 actual}
part_2(actual)
```


