# Lobby Layout

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(crayon.enabled = NULL, # for when rendering on github actions
        scipen = 999)          # make sure we never print in scientific notation
library(tidyverse)
```

This is my attempt to solve [Day 24](https://adventofcode.com/2020/day/24).

```{r load data}
sample <- read_lines("samples/day_24_sample.txt")
actual <- read_lines("inputs/day_24_input.txt")
```

## Part 1

First let's create a function that will iterate over the input and update a list of tiles. We can store the positions
of the tiles as complex numbers indicating the horizontal and vertical position of a tile.

If we move East or West we are moving exactly 1 unit right or left. If we move in one of the four North/South directions
we will instead move a half step right or left and a half step up or down.

```{r get tiles}
get_tiles <- function(input) {
  # each tile will be stored as 0 for white, 1 for black
  tiles <- list()
  # set the directions that we move
  directions <- c(
    e  =  1,
    se =  0.5 - 0.5i,
    sw = -0.5 - 0.5i,
    w  = -1,
    nw = -0.5 + 0.5i,
    ne =  0.5 + 0.5i
  )
  # iterate over the input, splitting each line into individual characters
  for (line in str_extract_all(input, ".")) {
    # we always start from (0, 0)
    p <- 0 + 0i
    # create an index into the current input line and iterate over it
    i <- 1
    while (i <= length(line)) {
      # get the current item in the line
      d <- line[[j]]
      # add in the next character if d is "s" or "n"
      if (d == "s" | d == "n") {
        i <- i + 1
        d <- paste0(d, line[[i]])
      }
      # update the current position
      p <- p + directions[[d]]
      i <- i + 1
    }
    # convert the position to a character so we can use as a name in the list
    p <- as.character(p)
    # if we haven't seen this tile yet, insert it as white
    if (is.null(tiles[[p]])) {
      tiles[[p]] <- 0
    }
    # flip the tile
    tiles[[p]] <- 1 - tiles[[p]]
  }
  # return the tiles
  tiles
}
```

We can use this function to solve part 1, we simply need to sum the results. But, because we return a list we first
need to flatten it to a double vector.

```{r part 1 function}
part_1 <- function(input) {
  input %>%
    get_tiles() %>%
    flatten_dbl() %>%
    sum()
}
```

We can test that this function works against the provided sample:

```{r part 1 sample test}
part_1(sample) == 10
```

And we can run this function against our actual data:

```{r part 1 actual}
part_1(actual)
```

## Part 2

```{r}
part_2 <- function(input, n) {
  directions <- c(
    e  =  1,
    se =  0.5 - 0.5i,
    sw = -0.5 - 0.5i,
    w  = -1,
    nw = -0.5 + 0.5i,
    ne =  0.5 + 0.5i
  )
  
  tiles <- get_tiles(input)
  
  for (i in 1:n) {
    # add in neighbours of black tiles
    missing_neighbours <- tiles %>%
      keep(~.x == 1) %>%
      names() %>%
      as.complex() %>%
      map(~ as.character(.x + directions)) %>%
      flatten_chr() %>%
      unique() %>%
      discard(~ .x %in% names(tiles))
    
    for (i in missing_neighbours) {
      tiles[[i]] <- 0
    }
    
    tiles <- imap(tiles, function(colour, position) {
      position <- as.complex(position)
      
      neighbours <- reduce(position + directions, .init = 0, function(s, n) {
        p <- as.character(n)
        s + ifelse(is.null(tiles[[p]]), 0, tiles[[p]])
      })
      
      if (colour == 0 && neighbours == 2) {
        1
      } else if (colour == 1 && (neighbours == 0 || neighbours > 2)) {
        0
      } else {
        colour
      }
    })
  }
  
  tiles %>%
    flatten_dbl() %>%
    sum()
}
```

We can test this function on the sample data:

```{r part 2 sample test}
part_2(sample, 100) == 2208
```

And we can run this function on the actual data:

```{r part 2 actual}
part_2(actual, 100)
```

Unfortunately this is terribly slow, taking nearly 9 minutes on my machine...
